### 2015-09-14 Manuscript/SW_early_life_telomeres_v9.Rmd
Updated figures and manuscript - standardised all response variables for within year analyses. Added markdown docx template (myref.docx)

---
output:
  word_document:
    fig_caption: yes
    highlight: null
bibliography: MyBib.bib
csl: JAE.csl
---
```{r load and clean, include = F}

#Set wd one level up
setwd('..')

#Load and clean data

source('Scripts/SW juvenile telomeres IPC load.r')
source('Scripts/SW juvenile telomeres IPC functions.r')
source('Scripts/SW juvenile telomeres IPC clean.r')

@@

There is evidence from a range of taxa that the greatest rate of telomere loss occurs in early life (e.g. [@Frenck1998; @Haussmann2003]), and that the extent of this early-life telomere shortening is influenced by the environment [@Boonekamp2014]. Moreover, a recent laboratory study showed that juvenile telomere dynamics are related to late-life survival [@Heidinger2012]. However, whether telomeres provide a link between early-life conditions and late-life survival in the wild is not known.

@@

In Seychelles warblers, oxidative stress is linked to natal territory quality [@VandeCrommenacker2011a], but neither natal territory quality nor natal local density are linked to juvenile or adult survival [@Brouwer2006; @Hammers2013]. In this study we analyse patterns of early-life telomere length and subsequent survival to test the prediction that telomeres fill a missing link between early-life environmental variation and late-life fitness. Specifically we predict that: i) individuals raised in good quality conditions will have longer telomeres, and ii) having longer telomeres in early life is associated with greater short and long-term survival.

@@

#####Study species and sampling
The Seychelles warbler is a small (~15 g) passerine bird with a mean life expectancy of 5.5 years at fledging [@Hammers2013]. The population of *ca.* 320 adults on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Richardson2003b; @Barrett2013a]. Individuals are usually ringed in their first year of life, and so are of known age. They are then followed throughout their lives, and as they are non-migratory endemics and are naturally confined to the island [@Komdeur2004a], a biannual census of birds on the island gives an accurate measure of levels of local density and individual survival rate [@Barrett2013a]. Full details of catching and census methods can be found in Brouwer *et al.* [@Brouwer2012].

All territories are mapped biannually based on detailed observational data of foraging and territorial defence behaviour, and surveyed for territory quality [@Richardson2003b]. Territory quality is calculated based on territory size, foliage cover and insect abundance [@Komdeur1992] and averaged within territories [@Hammers2013]. Cousin is also subject to inter-annual variation in rainfall and food availability, and such yearly effects may override the effects of absolute territory quality. As an estimate of annual variation in food availability, we calculated an index of the number of insects across the entire island during the main breeding season of each sampling year based on the insect abundance data (see above).

@@

Telomere lengths were measured from `r nrow(FlSA)+nrow(chicks)` samples taken cross-sectionally from birds aged less than twelve months, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`. Of these birds, `r nrow(chicks)` were nestlings less than two months old, `r nrow(subset(FlSA,Ageclass=='FL'))` were aged between two and six months (juveniles), and `r nrow(subset(FlSA,Ageclass=='SA'))` were aged seven to twelve months (subadults). We measured absolute telomere quantity per diploid genome using a quantitative PCR (qPCR) assay  (full details in [@Barrett2012]).

#####Statistical analyses
We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011]. We tested whether telomere length varied within the first year of life using one-way ANOVA, with telomere length as the response variable and age class (chick, juvenile or subadult) as a fixed factor. As telomere length varied substantially among age classes (see results) we carried out all subsequent analyses seperately for each age class.

We tested tested whether early-life environmental variation influences telomere length using linear mixed models, with telomere length as the response variable. As explanatory variables we included body condition (calculated as deviation from a regression line of mass on tarsus length; [@VandeCrommenacker2011a]), sex, territory quality, yearly food availability, and the number of helping and non-helping subordinates present in the natal territory  ([@Brouwer2012]). Lay Year was included as a random factor, in order to account for any annual variation in telomere length not explained by yearly food availability. We created a full model containing all of the above terms, and a top model set was then defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model averaged coefficients, confidence intervals, and 'relative importance', which reflects the relative weights of each predictor variable across the top model set. Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012].

We then used parametric survival analysis to examine variation in juvenile telomere length in relation to survival, implemented using the 'flexsurv' package in R. The survival time distribution was chosen by comparing the fit of a range of distributions (exponential, Weibull, log-normal, gamma, gompertz and generalized F) against the observed survival data, by visually inspecting plots and by comparing AIC values. As explanatory variables of survival we included juvenile telomere length, plus the factors identified as being related to telomere length in the previous analysis. With this approach we aimed to assess the effect of juvenile telomere length on survival while controlling for potentially confounding variables that may be correlated with telomere length.

For a subset of juveniles (n = `r nrow(Loss)` individuals) we had longitudinal data (i.e samples from individuals caught both as a juvenile an adult). For these individuals we calculated a rate of telomere loss between the juvenile sample and earliest adult sample by subtracting the adult telomere length from the juvenile telomere length and dividing this difference by the length of time (in days) between sampling events. To account for regression to the mean effects, we applied a correction based on correlations among samples within individuals following Verhulst *et al.* [-@Verhulst2013].

@@

Mean ($\pm$ standard error) early-life telomere length was `r round2(mean(FlSAall$TLKB))` $\pm$ `r round2(se(FlSAall$TLKB))` kb. This is only marginally higher than mean telomere length for all adult birds (`r round2(mean(subset(dd,Age>0)$TLKB))` $\pm$ `r round2(se(subset(dd,Age>0)$TLKB))` kb), and lower than the mean for one year-old birds based on cross-sectional data (`r round2(mean(subset(dd,Age==1)$TLKB))` $\pm$ `r round2(se(subset(dd,Age==1)$TLKB))` kb; Figure S1). However, when we considered ageclasses seperately, we found substantial age-related variation in telomere length within the first year of life (F = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,4])`; *P* = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,5])`; Fig. 1). This cross-sectional data revealed that chicks have the longest telomeres, but that subsadults and adults had longer telomeres than fledglings (Fig. 1). However, a longitudinal analysis showed that chicks, fleglings nd subadults all lost telomeres early in life, with no differences in rates of telomere shortening among age classes (F = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,5])`; Fig. S2). This suggests that the apparent increase in telomere length between juveniles and subadults in the corss-secitonal data was a result of the sampling, rather than telomere elongation.

In chicks, telomere length did not vary significantly among birth years (Fig. 2A, Fig. S2A).

Juvenile telomere length differed substantially among birth years (Figure S2), with longer telomeres in years where insect abundance was high (Figure 2B). Model averaging confirmed this strong, positive effect of yearly food availability on telomere length (Figure 2A; Table S1). The best supported model with juvenile telomere length as the response variable contained annual food availability as the only explanatory variable (Table S1; R^2^ = `r round2(summary(lm(TLKB~Insect,data=FlSAall))$r.squared)`).

A log-normal survival model best fitted the Seychelles warbler data (AIC = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`). A survival model including juvenile telomere length, body condition and early-life food availability showed that only juvenile telomere length had an effect on survival (estimate = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB+Insect), dist="lnorm")$res[3,1])`, CI = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB+Insect), dist="lnorm")$res[3,2])`-`r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB+Insect), dist="lnorm")$res[3,3])`; Fig. 2A). This survival effect was most pronounced in individuals between five and ten years of age (Fig. 2B). Because telomere length and annual food availability are highly correlated (*R* = `r round2(cor(FlSAall$Insect,FlSAall$TLKB)) #confidence test`), we ran an analysis of early-life telomere length on survival excluding early-life food availability, and the telomere effect beacme marginally non-significant (estimate = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,1])`, CI = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,2])`-`r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,3])`).

@@

###Authors' contributions
DSR developed the idea and obtained funding for the overall telomere project; DSR, JK, HLD and TB supervised the overall running of the overall Seychelles warbler project and contributed resources; HLD managed the longitudinal fitness database. EAF and KLB generated and processed the telomere data; LGS performed the analyses and wrote the manuscript in discussion with DSR. All authors read and contributed to the manuscript.

>>>>

---
output:
  word_document:
    fig_caption: yes
    highlight: null
bibliography: MyBib.bib
csl: JAE.csl
---
```{r setup, include = F}

rm(list=ls())

#Load knitr
library(knitr)

#Set some options
options(na.action='na.fail')
opts_knit$set(root.dir = '..')
opts_chunk$set(echo=FALSE,
               warning=FALSE,
               message=FALSE,
               fig.path='../Figures/',
               fig.width= 4,
               fig.height= 8,
               fig.show='asis',
               dev=c('png','pdf'))

```{r load and clean, include = F}

#Load and clean data
source('Scripts/SW juvenile telomeres IPC load.r')
source('Scripts/SW juvenile telomeres IPC functions.r')
source('Scripts/SW juvenile telomeres IPC clean.r')

@@

There is evidence from a range of taxa that the greatest rate of telomere loss occurs in early life (e.g. [@Frenck1998; @Haussmann2003]), and that the extent of this early-life telomere shortening is influenced by the environment [@Boonekamp2014]. Moreover, a recent laboratory study showed that early-life telomere dynamics are related to late-life survival [@Heidinger2012]. However, whether telomeres provide a link between early-life conditions and late-life survival in the wild is not known.

@@

In Seychelles warblers, oxidative stress is linked to natal territory quality [@VandeCrommenacker2011a], but neither natal territory quality nor natal local density are linked to early-life or adult survival [@Brouwer2006; @Hammers2013]. In this study we analyse patterns of early-life telomere length and subsequent survival to test the prediction that telomeres fill a missing link between early-life environmental variation and late-life fitness. Specifically we predict that: i) individuals raised in good quality conditions will have longer telomeres, and ii) having longer telomeres in early life is associated with greater short and long-term survival.

@@

#####Study species and sampling
The Seychelles warbler is a small (~15 g), insectivorous passerine bird with a mean life expectancy of 5.5 years at fledging [@Hammers2013]. The population on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Richardson2003b; @Barrett2013a]. Individuals are usually ringed in their first year of life, and so are of known age. They are then followed throughout their lives, and as they are non-migratory endemics and are naturally confined to the island [@Komdeur2004a], a biannual census of birds on Cousin during the main (June-August) and minor (December-February) breeding seasons gives an accurate measure of levels of local density and individual survival rate [@Barrett2013a]. Full details of catching and census methods can be found in Brouwer *et al.* [@Brouwer2012].

All territories are mapped during the main and minor breeding seasons based on detailed observational data of foraging and territorial defence behaviour, and surveyed for territory quality [@Richardson2003b]. Territory quality is calculated based on territory size, foliage cover and insect abundance [@Komdeur1992] and averaged within territories [@Hammers2013]. Cousin is also subject to intra- and inter-annual variation in rainfall and food availability, and such island-wide temporal variation may override the effects of absolute territory quality. As an estimate of variation in food availability, we calculated an index of the number of insects across the entire island during. This  index represents the average number of insects found per unit leaf area over all surveys carried out on the island in a breeding season.

@@

Telomere lengths were measured from `r nrow(FlSA)+nrow(chicks)` samples taken cross-sectionally from birds aged less than twelve months, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`. Of these birds, `r nrow(chicks)` were nestlings less than two months old (chicks), `r nrow(subset(FlSA,Ageclass=='FL'))` were aged between two and six months (fledglings), and `r nrow(subset(FlSA,Ageclass=='SA'))` were aged seven to twelve months (subadults). We measured absolute telomere quantity per diploid genome using a quantitative PCR (qPCR) assay  (full details in [@Barrett2012]).

#####Statistical analyses
We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011]. Our sampling regime covers a long time-period, and temporal variation in rainfall and food availability on Cousin may cause variation  in a range of social and evironmental variables [@Brouwer2012]. This makes it difficult to disentangle how spatial and temporal social and environmental variation affects early-life telomere dynamics. we seperately analysed how variation in early life conditions among and within breeding seasons was related to variation in early life telomere length.  First, we tested whether there was a temporal effect on telomere lengthby calculating average telomere length among all birds born in a given breeding season. We tested for a difference in amomg breeding seasons using a one-way ANOVA, and then used linear regression to test whether variation in mean telomere length among breeding seasons is explained by temporal variation in insect abundance.

We tested tested how spatial early-life environmental and social conditions influence telomere length within cohorts using linear mixed models. Telomere length was standarsised by mean-centering within cohorts (i.e. using all birds born in a given breeding season). As explanatory variables we included tarsus, length, sex, territory quality, yearly food availability, and the number of helping and non-helping subordinates present in the natal territory  ([@Brouwer2012]).

We used model averaging We created a full model containing all of the above terms, and a top model set was then defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model averaged coefficients, confidence intervals, and 'relative importance', which reflects the relative weights of each predictor variable across the top model set. Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012].

We used parametric survival analysis to examine variation in early-life telomere length in relation to survival, implemented using the 'flexsurv' package in R. The survival time distribution was chosen by comparing the fit of a range of distributions (exponential, Weibull, log-normal, gamma, gompertz and generalized F) against the observed survival data, by visually inspecting plots and by comparing AIC values. As explanatory variables of survival we included early-life telomere length, plus the factors identified as being related to telomere length in the previous analysis. With this approach we aimed to assess the effect of early-life telomere length on survival while controlling for potentially confounding variables that may be correlated with telomere length.

For a subset of first-year birds (n = `r nrow(Loss)` individuals) we had longitudinal data (i.e samples from individuals caught both in their first year and as an adult). For these individuals we calculated a rate of telomere loss between the first-year sample and earliest adult sample by subtracting the adult telomere length from the early-life telomere length and dividing this difference by the length of time (in days) between sampling events. To account for regression to the mean effects, we applied a correction based on correlations among samples within individuals following Verhulst *et al.* [-@Verhulst2013].

@@

#####Early-life telomere length varies with age and among breeding seasons
Mean ($\pm$ standard error) telomere length in first-year birds was `r round2(mean(FlSAall$TLKB))` $\pm$ `r round2(se(FlSAall$TLKB))` kb. This is lower than the mean for one year-old birds based on cross-sectional data (`r round2(mean(subset(dd,Age==1)$TLKB))` $\pm$ `r round2(se(subset(dd,Age==1)$TLKB))` kb; Figure S1). However, when we considered age classes seperately, we found substantial age-related variation in telomere length within the first year of life (F = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,4])`; *P* = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,5])`). This cross-sectional data revealed that chicks have longer telomeres than any other age group, but that subsadults and adults had longer telomeres than fledglings (Fig. 1A). A longitudinal analysis of telomere loss showed that chicks, fleglings and subadults all lost telomeres early in life, with highest and lowest rates of loss in chicks and subaldults, respectively, although the difference among age classes was not significant (F = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,5])`; Fig. 1B). This suggests that the apparent increase in telomere length between fledglings and subadults in the cross-sectional data was a result of the sampling, rather than telomere elongation.

Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; *P* = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`; Fig. 2A), and this variation in average telomere length over breeding seasons was positively related to island-wide food aviilability (linear regression, R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=juvseason))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason))[1,4])`; *P* = `r round2(anova(lm(Insect~TLKBmean,data = juvseason))[1,5])`; Fig. 2B). Similarly, the amount of telomeres lost in early life varied among breeding seasons (F = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,5])`; Fig. 2C), and individuals born in seasons with low food availability lost telomeres, on average, at a faster rate than those born in years with high food availability (linear regression, R^2^ = `r round2(summary(lm(Insect~TROCmean,data=loss.season))$r.squared)`; F = `r round2(anova(lm(Insect~TROCmean,data = loss.season))[1,4])`; *P* = `r round2(anova(lm(Insect~TROCmean,data = loss.season))[1,5])`; Fig. 2D).

#####Factors affecting early-life telomere dynamics within breeding seasons
The top explaining within-season variation in chick telomere length contained tarsus length, sex and the number of helpers present in the natal territory (Fig. 3A; Table S1). All three variables had high relative importance int he top model set (Fig. 3A), and the model containing these variables explained a reasonable amount of variation in telomere length (R^2^ = `r round2(summary(lm(TLKB~Tarsus+Helper+Sex,data=chickall))$r.squared)`). Female chicks had longer telomeres than males (Fig. 3B), while telomere length was positively related to the number of helpers (Fig. 3C), and negatively related to tarus to length (Fig. 3D). However, it should be noted that model-averaged confidence intervals for sex and the number of helpers (but not tarsus length) marginally overlapped zero. For both juveniles and sub-adults the null model (intercept only) was the top model (Table S1), and no explanatory terms had high relative importance in the top model set, with all estimates close to zero (Fig. 4).

For the subset of individuals with longitudinal data, we tested whether the factors previously identified as being related early-life telomere length (i.e. tarsus length, sex and number of helpers) were also associated with differences in rates of early life telomere loss. Due to a limited sample size, we considered all age classes together for this analysis. We found that none of these variables were associated with differences in telomere loss (all *P* > 0.05).

#####Early life telomere dynamics and survival
A log-normal survival model best fitted the Seychelles warbler data (AIC = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`). A survival model including juvenile telomere length, body condition and early-life food availability showed that only juvenile telomere length had an effect on survival (estimate = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB+Insect), dist="lnorm")$res[3,1])`, CI = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB+Insect), dist="lnorm")$res[3,2])`-`r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB+Insect), dist="lnorm")$res[3,3])`; Fig. 2A). This survival effect was most pronounced in individuals between five and ten years of age (Fig. 2B). Because telomere length and annual food availability are highly correlated (*R* = `r round2(cor(FlSAall$Insect,FlSAall$TLKB)) #confidence test`), we ran an analysis of early-life telomere length on survival excluding early-life food availability, and the telomere effect beacme marginally non-significant (estimate = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,1])`, CI = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,2])`-`r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,3])`).

@@

###Authors' contributions
DSR developed the idea and obtained funding for the overall telomere project; DSR, JK, HLD and TB supervised the overall running of the overall Seychelles warbler project and contributed resources; HLD managed the longitudinal fitness database. EAF, KLB and LGS generated and processed the telomere data; LGS performed the analyses and wrote the manuscript, in discussion with DSR. All authors read and contributed to the manuscript.
